COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-03 paper/bare_conf.tex
Refactor a few sections (msg of the paper targeted)

\author{\IEEEauthorblockN{Fabio Luporini\IEEEauthorrefmark{1},
Florian Rathgeber\IEEEauthorrefmark{1},
Ana Lucia Varbanescu\IEEEauthorrefmark{2}, 
Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
David Ham\IEEEauthorrefmark{1},
J. Ramanujam\IEEEauthorrefmark{3},
and
Paul H. J. Kelly\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
\IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$d.ham\}@imperial.ac.uk,\\
\IEEEauthorrefmark{2}a.l.varbanescu@uva.nl,
\IEEEauthorrefmark{3}jxr@ece.lsu.edu
}
}

@@

% make the title area
\maketitle

%The finite element method is widely-employed to approximate solutions of partial differential equations. Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation. Since the domain size can be huge, executing efficient kernels is fundamental. Automating their optimization is, however, a challenging issue. Even if an affine loop nest can be generally distinguished, small problem size and complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations. In this context, we present COFFEE, a domain-specific compiler for local assembly kernels. COFFEE manipulates kernels’ abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality. It then generates C code including vector intrinsics. Systematic performance evaluation using a suite of real-world examples shows that speed-ups over non-optimized kernels up to 4.44$\times$ are achievable.

\begin{abstract}
%\boldmath
The finite element method is widely-employed to approximate solutions of partial differential equations. Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation. Since the domain size can be huge, executing efficient kernels is fundamental. Automating their optimization is, however, a challenging issue. Even if an affine loop nest can be generally distinguished, small problem size and complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations. In this context, we present design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels. COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality. It then generates C code including vector intrinsics. Performance evaluation using a suite of real-world examples shows that signifcant speed-ups up are achievable.
\end{abstract}

@@

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs). Numerical techniques, like finite volume method and finite element method, are widely-employed to approximate solutions of PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each cell of the discretized domain by applying suitable numerical kernels. As the number of cells can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2isc}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.

@@

We focus on local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 30, with a minimum value of 3, depending on the order of the method). With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.

Optimization of assembly kernels is non-trivial. Given their structure and the exceptionally small size, assembly kernels benefit from transformations like generalized loop-invariant code motion, vector-register tiling and code splitting, which are not supported by state-of-the-art polyhedral and vendor compilers. BLAS routines could be theoretically employed, although fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies. BLAS libraries are also known to perform far from peak performance when the dimension of the matrices is small~\cite{nek5000}. As detailed in Section~\ref{sec:perf-results}, hand-made BLAS implementations of the Helmholtz assembly kernel (illustrated later) have run-times worse than those achieved with our optimization strategy.

%(our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, did not improve performance over $icc$ compilation)

%Also, tiling at the level of vector register, achievable through a non-affine transformation of the iteration space, has showed a significant performance improvement in a range of problems. .

@@

Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels. The contribution of this paper is threefold
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely local assembly in the context of the finite element method. Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g. $icc$) and research (e.g. polyhedral) compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements.
\end{itemize}

@@

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem like Burgers as in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the stiffness matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.

The class of kernels we are considering has, in particular, some peculiarities. 1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest. This is not true in general, since bent elements might be used in the unstructured mesh, which would require the Jacobian be re-computed at every $i$ iteration; 2) memory accesses along the three loop dimensions are always stride-1; 3) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 4) the $j$ and $k$ loops iterate over the same iteration space; 5) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

@@

In a local assembly kernel, although the bulk of the computation is carried out by the \texttt{ijk} affine loop nest, the typically small problem size and the complexity of the stiffness matrix evaluation make it hard to determine a single or unique sequence of successful transformations. Even though the memory access pattern is fixed, the fact that both data- and iteration-space sizes vary, possibly substantially, among different problems, render the optimization process challenging. Relatively to machine peak, while some problems achieve very good performance when applying a specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required. For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, will suffer from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} will not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of data and iteration spaces. Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this section, we provide a non-exhaustive list of transformations that local assembly kernels could benefit from. All of them are automated in COFFEE.

>>>>

\author{\IEEEauthorblockN{Fabio Luporini\IEEEauthorrefmark{1},
Florian Rathgeber\IEEEauthorrefmark{1},
Ana Lucia Varbanescu\IEEEauthorrefmark{2}, 
Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
David A. Ham\IEEEauthorrefmark{4},
J. Ramanujam\IEEEauthorrefmark{3},
and
Paul H. J. Kelly\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
\IEEEauthorrefmark{4}Imperial College London, Department of Computing and Department of Mathematics,\\
\IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$d.ham\}@imperial.ac.uk,\\
\IEEEauthorrefmark{2}a.l.varbanescu@uva.nl,
\IEEEauthorrefmark{3}jxr@ece.lsu.edu
}
}

@@

% make the title area
\maketitle
%OLD 1
%The finite element method is widely-employed to approximate solutions of partial differential equations. Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation. Since the domain size can be huge, executing efficient kernels is fundamental. Automating their optimization is, however, a challenging issue. Even if an affine loop nest can be generally distinguished, small problem size and complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations. In this context, we present COFFEE, a domain-specific compiler for local assembly kernels. COFFEE manipulates kernels’ abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality. It then generates C code including vector intrinsics. Systematic performance evaluation using a suite of real-world examples shows that speed-ups over non-optimized kernels up to 4.44$\times$ are achievable.

%OLD 2 before feedback
%The finite element method is widely-employed to approximate solutions of partial differential equations. Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation. Since the domain size can be huge, executing efficient kernels is fundamental. Automating their optimization is, however, a challenging issue. Even if an affine loop nest can be generally distinguished, small problem size and complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations. In this context, we present design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels. COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality. It then generates C code including vector intrinsics. Performance evaluation using a suite of real-world examples shows that signifcant speed-ups up are achievable.

\begin{abstract}
%\boldmath
Numerical solution of partial differential equations using the finite element method is one of the key applications of high performance computing. The local assembly is its characteristic operation, which executes a problem-specific kernel for each element in the discretized problem domain. Since the domain size can be huge, executing efficient kernels is fundamental. Their optimization is, however, a challenging issue. Even though affine loop nests are generally present, the short trip counts and the complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations. We present, therefore, the design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels. COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality. It then generates C code including vector intrinsics. Experiments using a suite of real-world examples show that significant performance improvement is achieved.
\end{abstract}

@@

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs). Numerical techniques, like finite volume method and finite element method, are widely-employed to approximate solutions of PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each element of the discretized domain by applying suitable numerical kernels. As the number of elements can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2isc}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.

@@

We focus on local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, whereas the minimum value is 3. In the innermost loop, a problem-specific expression evaluates a bi-dimensional array, representing the result of local assembly in an element of the discretized domain. With such a kernel structure, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.

Achieving high-performance is non-trivial. The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems. The variation in loop trip counts further exacerbates the issue. A compiler-based approach is, therefore, the only reasonable option to reach close-to-peak performance. Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but they are not supported by state-of-the-art polyhedral and vendor compilers. BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies. In addition, as detailed in Section~\ref{sec:perf-results}, in which manual BLAS-based implementations are evaluated, the small dimension of the involved matrices and loss in data locality can limit the performance gain, if any.

@@

Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels. The contribution of this paper is threefold
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely local assembly in the context of the finite element method. Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g. $icc$) and research (e.g. polyhedral) compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel machines, a Sandy Bridge and the Phi.
\end{itemize}

@@

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem like Burgers as in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the stiffness matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.

Some peculiarities of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

@@

As already emphasized, the structure of mathematical expressions evaluating the stiffness matrix and the variation in loop trip counts, although typically limited to the order of tens iterations, render the optimization process challenging. Relatively to machine peak, while some problems achieve good performance when applying a very specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required. For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces. Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.
